Data Processing: Mission Assignments

### 2021-11-11 code/processing_code/processing-combine.Rmd
EDA complete, linear analysis started

## Joining the Datasets
The two datasets could be joined by state or declaration number. 
```{r}
# #first change name of states column in missions df
# names(missions)[names(missions) == 'stateorTribe'] <- 'state'
# 
# #create df with states and declaration numbers from each df
# missions_nums <- as.data.frame(table(missions$state, missions$disasterNumber))
# declarations_nums <- as.data.frame(table(declarations$state, declarations$disasterNumber))

#we want a full join to keep all of the data
combined <- dplyr::full_join(declarations, missions, by = 'disasterNumber')
```

@@

It's not a flawless join, so let's do some investigation to see where there might be gaps.
```{r}
summary(combined)

#filter the rows that aren't in declarations data but are in missions data
missing_nums <- combined %>% 
                  filter(is.na(state)) %>%
                  select(disasterNumber)

#unique list of missing declarations
missing_declarations <- unique(missing_nums)

#check missing declarations in raw declarations
decs_check <- raw_decs %>%
                dplyr::filter(disasterNumber %in% missing_declarations$disasterNumber)

#the missing values represent declarations in territories, not states
#for now, remove - we can always come back and adjust as needed
combined_processed <- combined %>%
                        dplyr::filter(!is.na(IncidentYear))
```

@@

#save as an RDS
saveRDS(combined_processed, file = combined_data_location)
```

>>>>

## Preparing the datasets
The two datasets could be joined by state or declaration number. 
```{r}
# first change name of states column in missions df
names(missions)[names(missions) == 'stateorTribe'] <- 'state'

#find differences between disasterNumber and states included in each dataset
declarations_sub <- declarations %>% dplyr::select(disasterNumber,
                                                   state)

missions_sub <- missions %>% dplyr::select(disasterNumber,
                                           state)

missing_declarations <- dplyr::setdiff(missions_sub, declarations_sub)

#search in raw declarations
decs_check <- raw_decs %>%
                dplyr::filter(disasterNumber %in% missing_declarations$disasterNumber)
```
By examining the `decs_check` df, we can see that the missing declarations from the missions assignments do not exist in the declarations summaries. We can move forward and accept the reduction in data.

@@

It's not a flawless join, so let's do some investigation to see where there might be gaps.
```{r}
#use a left join to merge into the declarations df
combined <- dplyr::left_join(declarations, missions, by = c("state" = "state", "disasterNumber" = "disasterNumber"))

#the NAs for funding cost are declarations for which FEMA did not provide any funding
#therefore they can be replaced with 0
combined$TotalAgencies[is.na(combined$TotalAgencies)] <- 0
combined$ReqAmt[is.na(combined$ReqAmt)] <- 0
combined$OblAmt[is.na(combined$OblAmt)] <- 0
combined$FEMACSAvg[is.na(combined$FEMACSAvg)] <- 0

#summary of processed data
summary(combined)

```

@@

#save as an RDS
saveRDS(combined, file = combined_data_location)
```

### 2021-11-11 code/processing_code/processing-missions.Rmd
EDA complete, linear analysis started

Currently, the dataset is a mixture of wide and long formatting. To integrate effectively with the Disaster Declaration Summaries dataframe, we need to convert it completely to long formatting.
```{r long format}
#the gather function works positionally, so first reorder the columns to be in the right order
#essentially grouping columns by type of value (e.g. cost share percent, requested funding, obligated funding)
missions_6 <- missions_5[, c(1:3, 12, 6, 7, 4, 8, 9, 5, 10, 11)]

#trying piping several gather functions
missions_7 <- missions_6 %>%
                tidyr::gather(PayOrg, CostSharePct, TotalCSPct:FEMACSPct) %>%
                tidyr::gather(PayOrg2, ReqAmt, RequestedAmt:FEMAReqAmt) %>%
                tidyr::gather(PayOrg3, OblAmt, ObligatedAmt:FEMAOblAmt) %>%
                subset(select = -c(PayOrg2, PayOrg3))

#relabel values of `PayOrg`
missions_7$PayOrg[missions_7$PayOrg == "TotalCSPct"] <- "Total"
missions_7$PayOrg[missions_7$PayOrg == "StateCSPct"] <- "State"
missions_7$PayOrg[missions_7$PayOrg == "FEMACSPct"] <- "FEMA"
```

@@

After integrating the data, I have realized I really only care about the FEMA funds. So, just subset `PayOrg` to represent FEMA and then drop the `PayOrg` variable.
```{r FEMA filter}
missions_8 <- missions_7 %>%
                dplyr::filter(PayOrg == "FEMA") %>%
                subset(select = -c(PayOrg))

```

## Processed Data Overview
We have our processed dataset! Let's take a look at the overview to see how it changed.
```{r processed data overview}
#create a mirror set for documentation purposes
missions_processed <- missions_8

#summary using base R
summary(missions_processed)

#not using skimr because it does not do well with long format data
```

>>>>

Instead of trying to use agencies as a categorical variable, let's convert it to total number of agencies involved in the response. Then we can calculate the sum of the FEMA funding for each disaster number and each state
```{r long format}
#find count of distinct agencies for each disasterNumber and state
missions_6 <- missions_5 %>%
                dplyr::group_by(disasterNumber, stateorTribe) %>%
                dplyr::mutate(TotalAgencies = n_distinct(agency)) %>%
                subset(select = -c(agency))

#convert NaN to na for variables we're keeping
missions_6$FEMACSPct[is.nan(missions_6$FEMACSPct)] <- NA
missions_6$FEMAOblAmt[is.nan(missions_6$FEMAOblAmt)] <- NA
missions_6$FEMAReqAmt[is.nan(missions_6$FEMAReqAmt)] <- NA

#find sum of FEMA funds
missions_7 <- missions_6 %>%
                dplyr::group_by(disasterNumber, stateorTribe, TotalAgencies) %>%
                dplyr::summarize(ReqAmt = sum(FEMAReqAmt, na.rm = TRUE),
                                 OblAmt = sum(FEMAOblAmt, na.rm = TRUE),
                                 FEMACSAvg = mean(FEMACSPct, na.rm = TRUE))

#the NaN values in FEMACSAvg are when FEMA has no cost share (e.g. 0/0), so it can be replaced with 0
missions_7$FEMACSAvg[is.nan(missions_7$FEMACSAvg)] <- 0
```

@@

## Processed Data Overview
We have our processed dataset! Let's take a look at the overview to see how it changed.
```{r processed data overview}
#create a mirror set for documentation purposes
missions_processed <- missions_7

#summary using base R
summary(missions_processed)
```
